---
title: "Himalayan Expeditions"
author: "Erin Franke"
date: "2023-03-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(foreign)
```

```{r, message=FALSE, warning=FALSE}
expeditions <- read.dbf("HIMDATA/exped.DBF")
members <- read.dbf("HIMDATA/members.DBF")
peaks <- read.dbf("HIMDATA/peaks.DBF")
```

```{r}
members <- read.dbf("HIMDATA/members.DBF")
members <- members %>%
  filter(PEAKID %in% c("EVER", "AMAD", "CHOY", "MANA", "DHA1", "LHOT", "MAKA")) %>%
  filter(STATUS %in% c("Climber", "H-A Worker", "Leader"))
```

## Oxygen and success

Overall success

```{r}
members %>%
  count(MO2USED, MSUCCESS) %>%
  pivot_wider(id_cols = MSUCCESS, names_from = MO2USED, values_from =n) %>%
  rename("NoOxygen" = "FALSE", 
         "YesOxygen" = "TRUE") %>%
  mutate(percOxygen = round(YesOxygen/(YesOxygen+NoOxygen), 3))
```

## PEAK

Overall success: 44.55%

```{r}
members %>%
  count(MSUCCESS)
```

Success by peak:

```{r}
members %>%                                  
  group_by(MSUCCESS, PEAKID, MO2USED) %>%
  count() %>%
  pivot_wider(id_cols = PEAKID, names_from = MSUCCESS, values_from = n) %>%
  rename("Failure" = "FALSE", 
         "Success" = "TRUE") %>%
  mutate(percSuccess = round(Success/(Success+Failure), 3))
```

Using oxygen versus peak climbed

```{r}
members %>%
  group_by(PEAKID, MO2USED) %>%
  count() %>%
  pivot_wider(id_cols = PEAKID, names_from = MO2USED, values_from = n) %>%
  rename("NoOxygen" = "FALSE", 
         "YesOxygen" = "TRUE") %>%
  mutate(percOxygen = round(YesOxygen/(YesOxygen+NoOxygen), 3))
```

## Age versus success and oxygen used

Age doesn't really seem to impact if oxygen is used

```{r}
members %>%
  ggplot(aes(x=CALCAGE, fill = MO2USED))+
  geom_density(alpha=0.4)
```

Age might impact success a little bit

```{r}
members %>%
  ggplot(aes(x=CALCAGE, fill=MSUCCESS))+
  geom_density(alpha=0.4)
```

## Season

Oxygen is used far more often in spring than fall

```{r}
members %>%                                  
  group_by(MO2USED, MSEASON) %>%
  count() %>%
  pivot_wider(id_cols = MSEASON, names_from = MO2USED, values_from = n) %>%
  rename("NoOxygen" = "FALSE", 
         "YesOxygen" = "TRUE") %>%
  mutate(percOxygen = round(YesOxygen/(YesOxygen+NoOxygen), 3))
```

Success rates are highest in spring and fall and lowest in summer and winter

```{r}
members %>%                                  
  group_by(MSUCCESS, MSEASON) %>%
  count() %>%
  pivot_wider(id_cols = MSEASON, names_from = MSUCCESS, values_from = n) %>%
  rename("Failure" = "FALSE", 
         "Success" = "TRUE") %>%
  mutate(percSuccess = round(Success/(Success+Failure), 3))
```

## Sex

Oxygen use does not seem impacted by sex

```{r}
members %>%     
  filter(SEX != "X") %>%
  group_by(MO2USED, SEX) %>%
  count() %>%
  pivot_wider(id_cols = SEX, names_from = MO2USED, values_from = n) %>%
  rename("NoOxygen" = "FALSE", 
         "YesOxygen" = "TRUE") %>%
  mutate(percOxygen = round(YesOxygen/(YesOxygen+NoOxygen), 3))
```

Sex may impact success rate

```{r}
members %>%     
  filter(SEX != "X") %>%
  group_by(MSUCCESS, SEX) %>%
  count() %>%
  pivot_wider(id_cols = SEX, names_from = MSUCCESS, values_from = n) %>%
  rename("Failure" = "FALSE", 
         "Success" = "TRUE") %>%
  mutate(percSuccess = round(Success/(Success+Failure), 3))
```

```{r}
members %>%     
  filter(SEX != "X") %>%
  group_by(PEAKID, SEX) %>%
  count() %>%
  pivot_wider(id_cols = PEAKID, names_from = SEX, values_from = n) %>%
  mutate(percM = round(M/(M+F), 3))
```


## Leader

Leaders use oxygen at a lower rate than non-leaders

```{r}
members %>%     
  group_by(MO2USED, STATUS) %>%
  count() %>%
  pivot_wider(id_cols = STATUS, names_from = MO2USED, values_from = n) %>%
  rename("NoOxygen" = "FALSE", 
         "YesOxygen" = "TRUE") %>%
  mutate(percOxygen = round(YesOxygen/(YesOxygen+NoOxygen), 3))
```

Success doesn't necessarily seem dependent on whether you are a leader

```{r}
members %>%     
  group_by(MSUCCESS, STATUS) %>%
  count() %>%
  pivot_wider(id_cols = STATUS, names_from = MSUCCESS, values_from = n) %>%
  rename("Failure" = "FALSE", 
         "Success" = "TRUE") %>%
  mutate(percSuccess = round(Success/(Success+Failure), 3))
```

## INJURY

```{r}
members %>%     
  group_by(MSUCCESS, INJURY) %>%
  count() %>%
  pivot_wider(id_cols = INJURY, names_from = MSUCCESS, values_from = n) %>%
  rename("Failure" = "FALSE", 
         "Success" = "TRUE") %>%
  mutate(percSuccess = round(Success/(Success+Failure), 3))
```

```{r}
members %>%     
  group_by(MO2USED, INJURY) %>%
  count() %>%
  pivot_wider(id_cols = INJURY, names_from = MO2USED, values_from = n) %>%
  rename("Failure" = "FALSE", 
         "Success" = "TRUE") %>%
  mutate(percOxygen = round(Success/(Success+Failure), 3))
```

# IPW Exploratory Analysis

Should consider an interaction between season and peakid?

```{r}
 ggplot(members %>% filter(MSEASON %in% c("1", "3")), aes(x = factor(PEAKID), fill = factor(MO2USED))) +
    geom_bar(position = "fill") +
    facet_grid(~factor(MSEASON))

members %>% 
  filter(MSEASON %in% c("1", "3")) %>%
  group_by(MSEASON, PEAKID) %>%
  count() %>%
  pivot_wider(id_cols = PEAKID, names_from = MSEASON, values_from = n) %>%
  rename("Spring" = `1`, 
         "Fall" = `3`) %>%
  mutate(percSpring = round(Spring/(Spring+Fall), 3))
```

Don't need interactions between SEX/PEAKID or SEX/SEASON

```{r}
 ggplot(members %>% filter(SEX != "X"), aes(x = factor(PEAKID), fill = factor(MO2USED))) +
    geom_bar(position = "fill") +
    facet_grid(~factor(SEX))

 ggplot(members %>% filter(SEX != "X"), aes(x = factor(MSEASON), fill = factor(MO2USED))) +
    geom_bar(position = "fill") +
    facet_grid(~factor(SEX))
```


```{r}
members %>%
  ggplot(aes(x=CALCAGE, fill=STATUS))+
  geom_density(alpha=0.5)
```


Potentially line between status and sex

```{r}
members %>%   
  group_by(SEX, STATUS) %>%
  count() %>%
  pivot_wider(id_cols = STATUS, names_from = SEX, values_from = n) %>%
  mutate(percClimber = round(M/(M+`F`), 3))
```




